---
knit: bookdown::preview_chapter
---

### Examples

#### Weather in Ames 

Problem

What does today's temperature tell us about tomorrow's?

This question is inspired by Rob Hyndman's answer to it for temperatures in Melbourne as detailed in [@hyndman1996] and [@hyndman1996b].

Data source

The website Weather Underground [https://www.wunderground.com/](https://www.wunderground.com/)
provides current and historic weather information for weather stations all across the globe.
For historic data, each day is summarised by a set of statistics, such as temperature, humidity, pressure, and wind speeds and compared to averages. 

The process

We want to scrape the daily information from Weather Underground for a particular station and some  time frame. 

First, we write a scraper:
```{r}
# functions

getWeather <- function(station = "KAMW", year = 2015, month = 1, day = 1) {
  numbers <- "[-]*[0-9\\.]*[e0-9]*"
  units <- "[°a-zA-Z ]+"

  require(stringr)
  require(rvest)
  url <-
    sprintf(
      "https://www.wunderground.com/history/airport/%s/%s/%s/%s/DailyHistory.html", station, year, month, day
    )

  html <- read_html(url)
  tabs <- html %>%  html_nodes("table")
  day.stats <- html_table(tabs[[1]], fill = TRUE)

  if (is.null(day.stats$Actual)) return(NULL) # no data available
  day.stats$Unit <- str_extract(day.stats$Actual, pattern = units)
  day.stats$Actual <- as.numeric(str_extract(day.stats$Actual, pattern = numbers))
  day.stats$Average <- as.numeric(str_extract(day.stats$Average, pattern = numbers))
  day.stats$RecordYear <-
    str_extract(day.stats$Record, pattern = "\\([0-9]{4}\\)")
  day.stats$RecordYear <- gsub("[\\(\\)]","", day.stats$RecordYear)
  day.stats$Record <-
    as.numeric(str_extract(day.stats$Record, pattern = numbers))

  day.stats$Station <- station
  day.stats$Date <- sprintf("%s/%s/%s",year,month,day)
  names(day.stats)[1] <- "Statistics"

  day.stats
}
```

And now we scrape:
```{r, eval=FALSE}
######################
library(lubridate)

# download data for
last <- as.Date("2016-03-01")
first <- as.Date("2008-02-21")
days <- seq(first, last, "day")

filename <- "data/KAMW.csv"
station <- "KAMW"
for (dateIDX in length(days):1) {
#  browser()
  date <- days[dateIDX]
  res <- getWeather(station, year(date), month(date), mday(date))

  write.table(res, file=filename, append=file.exists(filename),
              col.names=!file.exists(filename),
              row.names=F, sep=",")
}
```

And now it's time for the analysis. Create a vector of the previous days temperatures and plot in a scatterplot. Alternatively, a two-dimensional histogram (using hex binning) also works well in this example, because the focus here is on the high density areas:
```{r, warning=FALSE}
weather <- read.csv("data/KAMW.csv", stringsAsFactors = FALSE)
weather$Actual <- as.numeric(weather$Actual)
weather$Average <- as.numeric(weather$Average)
weather$Record <- as.numeric(weather$Record)

temps <- subset(weather, Statistics=="Max Temperature")
temps$Previous <- c(temps$Actual[-1], NA)
library(ggplot2)
ggplot(data = temps, aes(x = Previous, y = Actual)) + 
  geom_jitter(alpha = I(0.5)) +  
  geom_smooth() + 
  xlab("Yesterday's maximum temperature") +
  ylab("Today's maximum temperature") + 
  coord_equal()

ggplot(data = temps, aes(x = Previous, y = Actual)) + 
  geom_hex() + 
  xlab("Yesterday's maximum temperature") +
  ylab("Today's maximum temperature") + 
  coord_equal()
```
What we see is a strong linear pattern, i.e. tomorrow's maximum temperature is close to what we see today, making it a very strong predictor. There is a peak in density at temperatures around 30 degree Celsius (shown in the binned scatterplot by the light blue tiles). With higher temperatures we see less variability, while at lower temperatures this relationship is more variable. 


#### Your turn

+ The pattern discussed involves maximum temperatures. How would you expect this pattern to change for averages? Explain your expectation first, then check. 

+ Reformat the weather data set in such a way that you can easily compare Maximum and minimum temperatures in a scatterplot. 

+ How do temperatures behave over the course of a year? Organize the data correspondingly and plot.

+ How does the temperature pattern compare to another place in the world, say e.g. Melbourne, Australia?


***Solution for last question***

Scrape the data
```{r, eval=FALSE}
last <- as.Date("2016-03-01")
first <- as.Date("1995-01-01")

last <- days[dateIDX]
first <- as.Date("1995-01-01")
days <- seq(first, last, "day")

filename <- "data/YMML.csv"
station <- "YMML"
for (dateIDX in length(days):1) {
#  browser()
  date <- days[dateIDX]
  res <- getWeather(station, year(date), month(date), mday(date))

  write.table(res, file=filename, append=file.exists(filename),
              col.names=!file.exists(filename),
              row.names=F, sep=",")
}

```
and plot the data:

```{r, warning=FALSE, eval=FALSE}
#weather <- read.csv("data/YMML.csv", stringsAsFactors = FALSE)
weather <- read_csv("data/YMML.csv")

temps <- subset(weather, Statistics=="Max Temperature")
temps$Previous <- c(temps$Actual[-1], NA)
ggplot(data=temps) +
  geom_hex(aes(x=Previous, y=Actual)) + #, fill=log(..value..))) + 
  xlab("Yesterday's maximum temperature") +
  ylab("Today's maximum temperature") +
  coord_equal() 
```

In the paper by [@hyndman1996b] conditional densities are used: given today's temperature what is the density distribution for tomorrow's temperatures?
We can use high-density region (HDR) boxplots to show the results. The location of the dot indicates the location of the density's highest mode.
```{r, fig.width=6, fig.height=6}
# bin yesterday's temperatures and compute densities of today's temperatures for each level
library(lubridate)
#temps$Month <- month(temps$Date)
#subtemps <- subset(temps, Month %in% c(12,1,2))
subtemps <- na.omit(temps[,c("Previous", "Actual")])

library(hdrcde)
temps.cde <- with(subtemps, cde(Previous, Actual, nxmargin=40))
plot(temps.cde,xlab="Yesterday's maximum temperature",ylab="Today's maximum temperature",plot.fn="hdr", prob=c(50, 95))
```

