---
knit: bookdown::preview_chapter
---

# Data Visualization


## A grammar of graphics

### What is a data plot?

- data
- **aesthetics: mapping of variables to graphical elements**
- geom: type of plot structure to use
- transformations: log scale, ...
- layers: multiple geoms, multiple data sets, annotation
- facets: show subsets in different plots
- themes: modifying style

### Why?

With the grammar, a data plot becomes a statistic. It is a functional mapping from variable to graphical element. Then we can do statistics on data plots. 

With a grammar, we don't have individual animals in the zoo, we have the genetic code that says how one plot is related to another plot.

### Elements of the grammar

```
ggplot(data = <DATA>) + 
  <GEOM_FUNCTION>(
     mapping = aes(<MAPPINGS>),
     stat = <STAT>, 
     position = <POSITION>
  ) +
  <COORDINATE_FUNCTION> +
  <FACET_FUNCTION>
```

The 7 key elements of the grammar are:

- DATA
- GEOM_FUNCTION
- MAPPINGS
- STAT
- POSITION
- COORDINATE_FUNCTION
- FACET_FUNCTION

## Example: Tuberculosis data

Data on tuberculosis incidence can be found at the [World Health Organisation](http://www.who.int/tb/country/data/download/en/). The case notifications table has been downloaded. We will look at the incidence in Australia. The steps to organising the data are:

1. Select the variables, country name and iso3 code, year and the incidence columns as measured by a sputum test. The incidence columns also encode age and gender.
2. Gather the data into long form, so that gender and age can be put in separate columns.
3. Filter out records for Australia, between years 1997-2012. We also remove values for under 15 because the coding of these age groups is contradictory. 

```{r}
library(tidyverse)
tb <- read_csv("data/TB_notifications_2018-03-18.csv") %>% 
  select(country, iso3, year, new_sp_m04:new_sp_fu) %>%
  gather(stuff, count, new_sp_m04:new_sp_fu) %>%
  separate(stuff, c("stuff1", "stuff2", "genderage")) %>%
  select(-stuff1, -stuff2) %>%
  mutate(gender=substr(genderage, 1, 1), 
         age=substr(genderage, 2, length(genderage))) %>%
  select(-genderage)

tb_au <- tb %>% 
  filter(country == "Australia") %>%
  filter(!(age %in% c("04", "014", "514", "u"))) %>%
  filter(year > 1996, year < 2013)
```

We want to look at this data with lots of plots in different ways, with different mappings, to learn different things. 

## 100% charts

```{r echo=TRUE, fig.width=8, fig.height=2}
ggplot(tb_au, aes(x = year, y = count, fill = gender)) +
  geom_bar(stat = "identity", position = "fill") +
  facet_grid(~ age) +
  scale_fill_brewer(palette="Dark2")
```

100% charts, is what excel names these beasts. What do we learn?

### Code structure

- Basic plot function is `ggplot`
- First argument provided is the name of the data, `tb_au`
- Variable mapping: year is mapped to x, count is mapped to y, gender is mapped to colour, and age is used to subset the data and make separate plots
- The bar geom is used, `geom_bar`
- We have already counted how many TB incidences are in each combination of categories, so `stat = "identity"` says no need to compute the count
- We are mostly interested in proportions between gender, over years, separately by age. The `position = "fill"` option in `geom_bar` sets the heights of the bars to be all at 100%. It ignores counts, and emphasizes the proportion of males and females.

### What do we learn?

- Focus is on **proportion** in each category. 
- Across (almost) all ages, and years, the proportion of males having TB is higher than females
- These proportions tend to be higher in the older age groups, for all years.


## Bar charts

```{r echo=TRUE, fig.width=8, fig.height=2}
ggplot(tb_au, aes(x = year, y = count, fill = gender)) +
  geom_bar(stat = "identity") +
  facet_grid(~ age) +
  scale_fill_brewer(palette="Dark2")
```

### What is different in the code description?

`, position = "fill"` was removed

### What do we learn?

- Focus is on **counts** in each category. 
- Different across ages, and years, counts tend to be lower in middle age (45-64)
- 1999 saw a bit of an outbreak, in most age groups, with numbers doubling or tripling other years.
- Incidence has been increasing among younger age groups in recent years.


## Side-by-side barcharts

```{r echo=TRUE, fig.width=8, fig.height=2}
ggplot(tb_au, aes(x = year, y = count, fill = gender)) +
  geom_bar(stat = "identity", position="dodge") +
  facet_grid(~ age) +
  scale_fill_brewer(palette="Dark2")
```

### What is different in the code description?

`, position="dodge"` is used in `geom_bar`

### What do we learn?

- Focus is on counts by gender, predominantly male incidence.
- Incidence among males relative to females is from middle ag on. There is similar incidence between males and females in younger age groups.

## Separate bar charts

```{r echo=TRUE, fig.width=8, fig.height=3}
ggplot(tb_au, aes(x = year, y = count, fill = gender)) +
  geom_bar(stat = "identity") +
  facet_grid(gender ~ age) +
  scale_fill_brewer(palette="Dark2")
```

### What is different in the code description?

`facet_grid(gender ~ age) +` faceted by gender as well as age, note `facet_grid` vs `facet_wrap`

### What do we learn?

- Its easier to focus separately on males and females.
- The 1999 outbreak mostly affected males.
- The growing incidence in the 25-34 age group is still affecting females but seems to be have stablised for males.

## Pie charts?

```{r echo=TRUE, fig.width=8, fig.height=3}
ggplot(tb_au, aes(x = year, y = count, fill = gender)) +
  geom_bar(stat = "identity") +
  facet_grid(gender ~ age) +
  scale_fill_brewer(palette="Dark2") + 
  coord_polar() +
  theme(axis.text = element_blank())
```

Nope! That's a rose chart. Bar charts in polar coordinates produce rose charts. 

### What is different in the code description?

`coord_polar() +` plot is made in polar coordinates, rather than the default Cartesian coordinates

### What do we learn?

- Emphasizes the middle years as low incidence. 

## Rainbow charts?

```{r echo=TRUE, fig.width=8, fig.height=3}
ggplot(tb_au, aes(x = 1, y = count, fill = factor(year))) +
  geom_bar(stat = "identity", position="fill") +
  facet_grid(gender ~ age) 
```

A single stacked bar, in each facet. Year is mapped to colour. 

### What is the code doing?

- Notice how the mappings are different. A single number is mapped to x, that makes a single stacked bar chart.
- year is now mapped to colour (that's what gives us the rainbow charts!)

### What do we learn?

- Pretty chart but not easy to interpret. 

## Pie charts

```{r echo=TRUE, fig.width=8, fig.height=3}
ggplot(tb_au, aes(x = 1, y = count, fill = factor(year))) +
  geom_bar(stat = "identity", position="fill") +
  facet_grid(gender ~ age) +
  coord_polar(theta="y") +
  theme(axis.text = element_blank())
```

### What is different in the code description?

`coord_polar(theta="y")` is using the y variable to do the angles for the polar coordinates to give a pie chart.

### What do we learn?

- Pretty chart but not easy to interpret, or make comparisons across age groups. 

### Your turn

Focus on a different country and work your way through the plots where we learned the most about the Australia data. 

1. Is the incidence similar for this country?
2. Are rates increasing? Or decreasing?
3. Do some age groups experience higher rates?
4. Is it more prevalent among males?